---
title: "training_data_renewed"
author: "Kevin"
date: "2025-02-23"
output: html_document
---

In this RMD file I will optimize my preprocessing of my  data for prediction of alpha-amylase content using spectral data.

Table Of Contents

1) Data Import and organization
2.) Noise Reduction 
3.) Data Transformation and Normalization
4.) Feature Extraction
5.) Data Augmentation and combination
6.) Final formatting

```{r}
# Import libraries

#Data Manipulation
library(dplyr) # Data manipulation
library(tidyr) # helps tidy Data
library(readr) # for data input/output
library(stringr) # Great for string manipulation
library(purrr) # supports working with vectors and lists
library(reshape2) # Data reshaping
library(tidyverse)

# Data Visualization packages
library(ggplot2)
library(plotly)
library(cowplot)
library(RColorBrewer) # for color palettes

#Spectral Analysis 
library(hyperSpec) # process, analyze, visualize hyperspec data
library(prospectr) # can be used for its preprocessing functions
library(signal) # Used for Savitzy-Golay 
library(mdatools) # Multivariate data analysis
library(mvoutlier) # outlier detection
library(lightr)
library(baseline) # Used for Baseline correction


#Machine Learning and Modeling
library(nnet)
library(caret)

#Timeseries
library(zoo) # allows us to fill NA
```


1.) Import and organization
```{r}
# Read the raw AA data from a CSV file. Also Try to use this for adding any new training data.
data_april_25 <- read.csv("/Users/propst/Desktop/using/april_25_seed_spectral_data.csv", header = TRUE)

```


```{r}

# Remove X from in front of wavelength.
colnames(data) <- sub("^X", "", colnames(data))

# Individual seed. Also make sure that it in the germination stage.

# Change wavelength to numeric 
data_july <- na.omit(data)

# Individual seed
data_individual <- data_july%>%
 dplyr::filter(as.numeric(seed_num) >0 & as.numeric(seed_num) < 49,
        side == "v",
        stage == "g")
#Remove .raw from date column.
data_individual$date <- gsub("\\.raw", "", data_individual$date)

```
2.) Baseline Correction 
```{r}

# 1. Identify spectral columns
spectral_cols <- grep("^[0-9]+\\.[0-9]+$", names(data_individual), value = TRUE)

# 2. Extract spectral data and convert to matrix
spec_matrix <- data_individual %>%
  select(all_of(spectral_cols)) %>%
  as.matrix()

# 3. Apply baseline correction using the ALS method
bl_result <- baseline.als(spec_matrix, lambda = 50, p = 0.1)

# 4. Extract the corrected spectra from the result
corrected_spec <- bl_result$corrected

# 5. Combine the corrected spectral data back with the original metadata
data_individual_corrected <- data_individual %>%
  select(-all_of(spectral_cols)) %>%       # Remove original spectral columns
  bind_cols(as.data.frame(corrected_spec))   # Bind corrected spectra

# Optional: inspect the corrected data
head(data_individual_corrected)



# Reshape the corrected data from wide to long format
df_plot <- data_individual_corrected %>%
  pivot_longer(
    cols = matches("^[0-9]+\\.[0-9]+$"),  # select columns with wavelength names
    names_to = "wavelength",
    values_to = "reflectance"
  ) %>%
  mutate(wavelength = as.numeric(wavelength))

df_plot <- df_plot %>%
  dplyr::filter(wavelength > 420)
# Create the plot: facet by timepoint
ggplot(df_plot, aes(x = wavelength, y = reflectance, group = seed_num, color = genotype)) +
  geom_line(alpha = 0.7) +
  facet_wrap(~ timepoint, scales = "free_y") +
  labs(title = "Baseline Corrected Spectra by Timepoint",
       x = "Wavelength (nm)",
       y = "Corrected Reflectance") +
  theme_minimal()
```
3.) Noise reduction


```{r pressure, echo=FALSE}

# Apply Savitzky-Golay Filter
df_long_golay <- df_plot %>%
  group_by(side, stage, date, timepoint, genotype, seed_num) %>%
  arrange(wavelength) %>%  # Ensure wavelengths are in increasing order
  mutate(smoothed_reflectance = sgolayfilt(reflectance, p = 2, n = 11)) %>% 
  ungroup()

head(df_long_golay)

# Create a comparison plot with raw and smoothed reflectance
ggplot(df_long_golay, aes(x = wavelength)) +
  geom_line(aes(y = reflectance, color = "Raw"), size = 1, linetype = "dashed") +
  geom_line(aes(y = smoothed_reflectance, color = "Smoothed"), size = 1) +
  labs(title = "Comparison of Raw vs. Savitzky-Golay Smoothed Spectra",
       x = "Wavelength",
       y = "Reflectance",
       color = "Data Type") +
  theme_minimal()
```
4.) Data Tranformation
```{r}
library(signal)

# 1) Frist Derivative
df_long_deriv <- df_long_golay %>%
  group_by(side, stage, date, timepoint, genotype, seed_num) %>%
  arrange(wavelength) %>%  
  # Compute first derivative of the smoothed reflectance
  mutate(first_derivative = sgolayfilt(smoothed_reflectance, p = 2, n = 11, m = 1)) %>%
  ungroup()

# Inspect a portion of the transformed data
head(df_long_deriv)

# 2.) Standard Normal Variate (SNV)
df_long_snv <- df_long_golay %>%
  group_by(side, stage, date, timepoint, genotype, seed_num) %>%
  arrange(wavelength) %>%
  # Compute SNV for each spectrum within each group
  mutate(snv_reflectance = (smoothed_reflectance - mean(smoothed_reflectance)) / sd(smoothed_reflectance)) %>%
  ungroup()

# Inspect a portion of the SNV-transformed data
head(df_long_snv)

# Assume 'spectral_cols' contains the names of the spectral columns, and
# data_individual_corrected is your data frame with baseline‐corrected and smoothed spectra

# Extract spectral data matrix:
spec_matrix <- df_long_snv %>%
  select(all_of(spectral_cols)) %>%
  as.matrix()

# Create a numeric vector of wavelengths from your column names:
wavelength <- as.numeric(spectral_cols)
spec_msc <- msc(spec_matrix)

library(ggplot2)

# Choose a sample index to inspect (e.g., first spectrum)
spec_index <- 1

# Create a data frame for plotting
df_plot <- data.frame(
  wavelength = wavelength,
  original = spec_matrix[spec_index, ],
  msc_corrected = spec_msc[spec_index, ]
)

# Plot the original and MSC-corrected spectra
ggplot(df_plot, aes(x = wavelength)) +
  geom_line(aes(y = original, color = "Original")) +
  geom_line(aes(y = msc_corrected, color = "MSC Corrected")) +
  labs(title = "Spectrum Comparison",
       x = "Wavelength",
       y = "Reflectance") +
  scale_color_manual(values = c("Original" = "black", "MSC Corrected" = "red"))

```

3.) Normalization

```{r}

# Create Normalization function
min_max_normalize <- function(x) {
  return ((x - min(x)) / (max(x) - min(x)))
}

# Apply function to filtered data
df_long_normal <- df_long_deriv %>%
  group_by(timepoint) %>%
  mutate(norm_reflectance = min_max_normalize(reflectance))

df_long_normalized_smooth <- df_long_deriv %>%
  group_by(timepoint) %>%
  mutate(norm_reflectance_smooth = min_max_normalize(smoothed_reflectance))

df_long_normalized_deriv <- df_long_deriv %>%
  group_by(timepoint) %>%
  mutate(norm_reflectance_deriv = min_max_normalize(first_derivative))
```

```{r}
library(ggplot2)
library(dplyr)

# Plot 1: Normalized Raw Reflectance
p1 <- ggplot(df_long_normal, aes(x = wavelength, y = norm_reflectance)) +
  geom_line(color = "blue") +
  labs(title = "Min-Max Normalized Raw Reflectance",
       x = "Wavelength", y = "Normalized Reflectance") +
  facet_wrap(~ timepoint, scales = "free_y") +
  theme_minimal()

# Plot 2: Normalized Smoothed Reflectance
p2 <- ggplot(df_long_normalized_smooth, aes(x = wavelength, y = norm_reflectance_smooth)) +
  geom_line(color = "darkgreen") +
  labs(title = "Min-Max Normalized Smoothed Reflectance",
       x = "Wavelength", y = "Normalized Smoothed Reflectance") +
  facet_wrap(~ timepoint, scales = "free_y") +
  theme_minimal()

# Plot 3: Normalized First Derivative

df_long_normalized_deriv <- df_long_normalized_deriv%>%
  dplyr::filter(wavelength >450 & wavelength <1000)
p3 <- ggplot(df_long_normalized_deriv, aes(x = wavelength, y = norm_reflectance_deriv)) +
  geom_line(color = "red") +
  labs(title = "Min-Max Normalized First Derivative",
       x = "Wavelength", y = "Normalized First Derivative") +
  facet_wrap(~ timepoint, scales = "free_y") +
  theme_minimal()

p1
p2
p3
```



